Computing MEM seeding probabilities

### 2019-02-07 MEM.tex
Added the part on false positives.

\title{Calibrating MEM seeding heuristics}

@@

The focus of this document is solve the true location problem \textit{per
se}, but rather to evaluate whether a solution is correct. At first sight,
whether we can map reads to their true location seems to depend only on
their lengths and on the error rate of the instrument. However, there is
another important factor to consider: most genomes contain repeated
sequences, \textit{i.e.} relatively large subsequences (from $\sim20$~bp
to more than 10~kb) that are present at more than one location. So we
cannot map the fragment with certainty if there exists an exact copy of
the sequence at somewhere else in the genome, because we cannot know which
of the two copies was sequenced.

@@

\subsection{Seeding heuristics}

@@

This definition presents a computational challenge: to know if a given
subsequence is a MEM seed, we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to identify all the MEM seeds from a read, even
for very large genomes\cite{ferragina2000opportunistic,
ferragina2005indexing}. Such algorithms are crucial for the present
discussion, but describing them falls outside the scope of the document.
We refer the interested reader to the relevant
literature\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
li2013aligning}, and here on we will assume that MEM seeds are simply
available at all times.

@@

@@

\section{Combinatorial construction of reads}

In this section we represent reads as combinatorial ojects. The
construction has two essential purposes: the first is to give a generative
model of reads that do not have any on-target MEM seed; the second is to
allow us to find the weighted generating function of those reads, which in
turn will allow us to find their probability of occurrence.

@@

The nucleotide sequence of the read is inadequate to know where MEM seeds
are and whether they are on target. For this purpose, we will recode reads
as sequences of letters from the four letter \emph{MEM alphabet}
$\{\downarrow, \uparrow, |, \square\}$.

Each nucleotide is recoded with $\downarrow$, $\uparrow$, or $\square$.
The $\downarrow$ symbol indicates that the nucleotide is a sequencing
error. By definition, it is a mismatch for the target. The $\uparrow$
symbol indicates that the nucleotide is the ``first'' nucleotide of an
on-target MEM seed, which we are about to clarify. All the other
nucleotides are represented by the $\square$ symbol. Finally, the $|$
symbol is appended after the last nucleotide of the read. It is an
obligatory terminator that does not correspond to an actual nucleotide.

To explain the precise meaning of the $\uparrow$ symbol, let us consider
the left end of the read: as we try to extend the first match as much as
possible to the right, the number of hits in the genome decreases. It may
be the case that at some point, there is a single hit and that it is on
target. The very first nucleotid where that happens is called the
\emph{pivot} and is labelled with the symbol $\uparrow$. An on-target MEM
seed has exactly one pivot, so the number of $\uparrow$ symbols is the
number of on-target MEM seeds in the read. Note that $\uparrow$ symbols
marks all on-target MEM seeds, regardless of their size.

Fig.~\ref{fig:sketch_MEM} shows the read of Fig.~\ref{fig:MEM_example} as
encoded with the MEM alphabet. Below the read are the matches between the
read and the potential locations from Fig.~\ref{fig:MEM_example}. A black
square means that the nucleotide is a mismatch for the genomic sequence,
and a white or grey square means that it is a match. Grey squares are used
to highlight MEM seeds, as in Fig.~\ref{fig:MEM_example}.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_MEM.pdf}
\caption{\textbf{The MEM encoding}.
The read of Fig.~\ref{fig:MEM_example} is recoded with the MEM alphabet.
The mismatches between the read and the potential locations are indicated
with black boxes and the matches with white boxes. The grey boxes are
matches and they indicate the locations of the MEM seeds. The $\downarrow$
symbols coincide exactly with the mismatches between the read and the
target. The $\uparrow$ symbol is always a match for the target and a
mismatch for some off-target location. It indicates the presence of a
strict on-target MEM seed (flanked by the nearest $\downarrow$ symbols or
by the end of the read). Also note the presence of the $|$ terminator at
the end of the read.}
\label{fig:sketch_MEM}
\end{figure}

Observe that the two on-target MEM seeds contain a pivot, but the shared
MEM seed does not. This is always the case. Recall that the pivot
indicates that the true location is the \emph{only} local match, so the
$\uparrow$ symbol cannot appear in on-target MEM seeds when they are
shared.

The MEM alphabet makes it easy to find strict on-target MEM seeds: they
are the longest stretches of symbols containing $\uparrow$ and not
containing $\downarrow$. However, shared on-target MEM seeds do not appear
in this representation.

\subsection{The extended MEM alphabet}

To make shared on-target MEM seeds apparent, we decorate the $\uparrow$
and $\downarrow$ symbols, effectively extending the alphabet. The
$\downarrow$ symbol is replaced by the set of symbols $\downarrow_{/m}$,
where $m$ is the number of duplicates that \emph{match} the erroneous
nucleotide. The $\uparrow$ symbol is replaced by the set of symbols
$\uparrow_{/j}$, where $j-1$ is the number of consecutive $\square$
symbols preceding $\uparrow$.

Note that the numbers decorating the symbols have a very different nature.
In the case of the $\downarrow$ symbol, it is an information about the
number of sequences matching the read; in the case of the $\uparrow$
symbol, it is the position of the pivot relative to the previous
sequencing error (or relative to the beginning of the read).

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_extended.pdf}
\caption{\textbf{The extended MEM encoding}.
The read of Fig.~\ref{fig:sketch_MEM} is represented in the extended MEM
alphabet. The symbols $\uparrow$ and $\downarrow$ are now decorated by
numbers. The arrows departing from those numbers help understand their
meaning. The $\downarrow$ symbol is decorated by the number of locations
that match the nucleotide. The $\uparrow$ symbol is decorated by the
number of consecutive $\square$ symbols before it, plus one.}
\label{fig:sketch_extended}
\end{figure}

Fig.~\ref{fig:sketch_extended} shows the encoding of the read from
Fig.~\ref{fig:sketch_MEM} in the extended MEM alphabet. The arrows bring
the focus on how the numbers decorating the $\uparrow$ and $\downarrow$
symbols are computed. For instance, the $\downarrow_{/0}$ symbol indicates
that the nucleotide is an error (this is the meaning of the $\downarrow$
symbol), and that it matches no other location (this is meaning of $/0)$.
In other words, it is a mismatch against every potential location.

In this new encoding, shared on-target MEM seeds are stretches of
$\square$ symbols flanked either by $\downarrow_{/0}$, or by $|$, or by
the left end of the read. Indeed, such a stretch is a MEM seed because it
matches at least one sequence (the target) and it cannot be extended
(because the end of the read is reached or because there is no match in
the genome). Also, it cannot be a strict on-target MEM seed because it
does not contain the $\uparrow$ symbol, so it must be a shared on-target
MEM seed.

The reason for introducing more than just the $\downarrow_{/0}$ symbol is
not immediately clear. For the combinatorial construction to be useful, we
need not only to identify on-target MEM seeds, but also to compute the
probabilities of the reads of each type. This is precisely the role of the
extended alphabet, as we will see in section~\ref{sec:transfer_mat}.

@@

The final twist in the combinatorial construction is to consider reads not
as sequences of symbols, but as sequences of \emph{segments}. The purpose
here is merely to express that some on-target MEM seeds are good
(\textit{i.e.}, their size is $\gamma$ or greater) and some are not.

@@

In line with definition~\ref{def:segment}, the symbols $\uparrow_{/j}$,
$\downarrow_{/m}$ and $|$ are referrred to as \emph{terminators}.
Fig.~\ref{fig:sketch_segment} shows the segment representation of the read
from Fig.~\ref{fig:sketch_extended}. The numbers below indicate the size
of each segment.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_segments.pdf}
\caption{\textbf{The extended MEM segment encoding}.
The read from Fig.~\ref{fig:sketch_segment} is represented as a sequence
of segments. Each segment is terminated by either $\uparrow_{/j}$, or
$\downarrow_{/m}$ or $|$. The rightmost fragment is the tail. The numbers
below indicate the size of each segment in nucleotides.}
\label{fig:sketch_segment}
\end{figure}

Shared on-target MEM seeds at the tail of the read are segments but all
other on-target MEM seeds are not. Shared MEM seeds cannot contain the
terminator $\downarrow_{/0}$ because it is a mismatch against the target.
Strict on-target MEM seeds are split over at least two segments: one
terminated by $\uparrow_{/j}$ and one right of it terminated by either
$\downarrow_{/m}$ or $|$.

\section{Symbolic and analytic representations}

We now turn to the question of translating the extended MEM segment
representation to a construction of reads without good seeds
(\textit{i.e.}, without on-target MEM seeds of size $\gamma$ or greater).
For this, we use tools borrowed from analytic combinatorics. We give the
minimum amount of background that is necessary to expose the theory. A
more detailed account of the strategy to compute seeding probabilities
with analytic combinatorics has been published
elsewhere\cite{filion2017analytic,filion2018analytic}. We refer the
readers interested in learning more about analytic combinatorics to the
textbooks\cite{flajolet2009analytic,sedgewick2013introduction}.

@@

Let $\mathcal{A}$ be a set of combinatorial objects with a size $k \in
\mathbb{N}$ and a probability of occurrence. Denoting $a_k$ the total
probability of objects of size $k$, the function defined as $A(z) =
\sum_{k\geq0} a_kz^k$ is called the \emph{weighted generating
function} of the combinatorial objects.

In our case, we are interested in the weighted generating function of
reads without good seeds, so the coefficient $a_k$ is the probability
that a read of size $k$ has no good seed. As $k$ becomes larger,
it becomes more difficult to evaluate $a_k$, but the representation
developed above will allow us to do this.

\subsection{The symbolic method}

@@

For each potential location including the target, we define the
\emph{match streak} at a given nucleotide as the number of nucleotides
since the last mismatch on the left. For instance, in
Fig.~\ref{fig:MEM_example}, the match streak of the target at the ninth
nucleotide is $1$ while for the second potential location, it is $7$.

For instance, the second location in the example above is a hard mask at
the ninth nucleotide. On Fig.~\ref{fig:MEM_example}, one can further check
that all the off-target locations are soft masks at the second nucleotide
and that the second location is a soft mask at the last nucleotide of the
read.

A nucleotide with a hard mask cannot be part of an on-target MEM seed. A
nucleotide without mask belongs to a strict on-target MEM seed. A
nucleotide with a soft mask may belong to a strict on-target MEM seed, to
a shared on-target MEM seed, or may not belong to any MEM. This case is
the most intricate and it will be teased apart below.

The extended MEM alphabet is nothing more than a way to keep track of the
hard and soft masks. For instance, a pivot (associated with the symbols
$\uparrow_{/j}$) is a nucleotide where the number of masks becomes 0.
Denoting the number of off-target locations by $N$, the symbol
$\downarrow_{/m}$ indicates that there are $m$ hard masks and $N-m$ soft
masks.

@@

We assume that the target sequence has $N$ duplicates. We further assume
that duplication was instantaenous and that all $N+1$ sequences diverge
independently from each other at a constant rate. In other words, we
ignore the complications due to the genealogy of the duplication events
and we simply assume that at each position, any given duplicate is
identical to the target with probability $1-\mu$. If it is not, we
assume that the duplicate has any of the remaining three nucleotides with
equal probability (\textit{i.e.} each is found with probability $\mu/3$).

@@

On a read error the target is never identical to the read (because we
assume that the target is the true sequence), and a duplicate is identical
to the read with probability $mu/3$ (the probability that the duplicate is
different from the target, times the probability that the nucleotide is
the same as the error). On a correct nucleotide, the target is always
identical to the read, and a dupicate is identical with probability
$1-\mu$ (the probablity that the duplicate is identical to the target).

With these assumptions, we are now ready to describe the entries of the
transfer matrix $M(z)$, which we do in order of increasing difficulty.

@@

A $\uparrow_{/j}$ symbol marks a pivot and the absence of masks. The
segment is thus the beginning of an on-target MEM of size at least $j$.
The MEM will extend until the next sequencing error, or until the end of
the read, so the number of $\square$ symbols in the next segment must be
at most $\gamma-j-1$ and it must be terminated by a $\downarrow$ symbol or
by the tail terminator $|$.

In this case, the matches between the read and the duplicates are
irrelevant until the terminator. The weighted genearting function of each
$\square$ symbol is simply $qz$ ($q$ is the probability that the
nucleotide is not an error and $z$ marks objects of size $1$).

@@

The weighted generating function of the $\downarrow_{/m}$ terminator is
thus $\omega_m pz$ (the probability of the $\downarrow_{/m}$ is $\omega_m
p$ and $z$ marks objects of size $1$).

@@

At a $\downarrow_{/m}$ symbol, there are $m$ hard masks and $N-m$ soft
masks. If all the masks disappear before the first read error, the next
terminator will be a $\uparrow$ symbol, otherwise it will be a
$\downarrow$ symbol.

@@

\begin{definition}
The probability that a given duplicate sequence contains a mismatch in a
sequence of $i$ error-free nucleotides is
\begin{equation}
\label{eq:xi}
\xi_i = 1-(1-\mu)^i.
\end{equation}
\end{definition}

@@

In the equation above, $q^i$ is the probability that there is no
sequencing error in $i$ nucleotides and $z^i$ marks objects of size $i$.
As mentioned above, $i < \gamma$ for reads without on-target MEM seed.
Also note that this expression is the same for all $\downarrow$ symbols;
it does not depend on $m$.

@@

In this case there can be no on-target MEM seed. We must just enforce the
condition that at least one of the $N$ masks survives until the end of the
segment in order to exclude the $\uparrow$ symbol. The weighted generating
function is
\begin{equation*}
\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i.
\end{equation*}

@@

In this case, the soft masks are irrelevant. Even if they survive until
the end of the segment, there will be a shared on-target MEM seed. So we
must enforce the condition that at least one hard mask survives until the
end of the segment (which is impossible if $m = 0$).
\begin{equation*}
\sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
\end{equation*}

The transition from $\downarrow_{/m}$ to $|$ is
\begin{equation}
\label{eq:C}
C_m(z) =
\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i +
  \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
\end{equation}

@@

As above, there can be no on-target MEM seed and we must just exclude the
$\uparrow$ terminator. The weighted generating function is
\begin{equation}
\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i.
\end{equation}

\subsubsection*{Case 3a: the terminator $\downarrow_{/n}$ comes after the
$\gamma$-th nucleotide}

This case is by far the most convoluted. Since the segment contains at
least $\gamma$ error-free nucleotides, we must take care that it does not
generate an on-target MEM seed. This will be the case if any of the two
following conditions is validated: $i)$ at least one hard mask covers all
the error-free nucleotides, or $ii)$ none of the hard masks does but at
least one soft mask covers the whole segment (including the terminator).

The two conditions are mutually exclusive by construction. They are
graphically represented in the diagram below. The left panel corresponds
to case $i)$ and the panel to case $ii)$. The top row represents the
target, and the bottom rows represent duplicates (using the same
symbols as in Fig.~\ref{fig:sketch_MEM}).
\begin{inset}
\includegraphics{masks.pdf}
\end{inset}

Whenever a hard mask (here the first duplicate) covers the nucleotides
as shown in the left panel, there can be no on-target MEM seed,
irrespective of the value of $\gamma$. The positions marked with a
question mark are irrelevant, they cannot change the fact that there is no
on-target MEM seed. If the hard masks are interrupted, as in the right
panel, then it depends on the soft masks. If a soft mask covers the whole
segment, then there can be no on-target MEM seed, irrespective of the
value of $\gamma$.

@@

For condition $ii)$, we will need to introduce some new notations.
\begin{definition}
The probability that a given duplicate sequence contains a mismatch in a
sequence of $i$ error-free nucleotides followed by an error is
\begin{equation}
\label{eq:eta}
\eta_i = 1-(1-\mu)^i\mu/3.
\end{equation}
\end{definition}

The probability of condition $ii)$ is
\begin{equation*}
\xi_i^m \Big(1 - \eta_i^{N-m} \Big),
\end{equation*}
but we need to break up this term among the terminators $\downarrow_{/n}$,
($0 \leq n \leq N$). For this, we decompose the sum on the number of soft
masks that run to the end of the segment, including the terminator. The
probability that there are $r \geq 1$ such soft masks is
\begin{equation*}
{N-m \choose r} (1 - \eta_i)^r \eta_i^{N-m-r}.
\end{equation*}

Each of them matches the nucleotide at the end of the segment, so the
total number of matches is $r$ plus the number of sequences, among the
remaining $N-m-r$ soft masks and $m$ hard masks, that also match the
last nucleotide.

Let us start with the $N-m-r$ that failed somewhere in the segment. The
conditional probability that they match the last nucleotide given that
they fail somewhere in the segment is $\mu/3 \cdot \xi_i / \eta_i$. The
conditional probability that they do not match is $(1-\mu/3) / \eta_i$.
For the $m$ hard masks, the probability that they match the last
nucleotide given that they sail somewhere in the segment is simply
$\mu/3$, and the probability that they do not is $1-\mu/3$.

The probability that those $N-r$ duplicates contribute $n-r$ matches is
\begin{equation*}
\frac{(\mu/3)^{n-r}(1-\mu/3)^{N-n}}{\eta_i^{N-m-r}}
\psi_{i,m,n,r}\text{, where \;}
\psi_{i,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
\xi_i^{n-r-q}.
\end{equation*}

>>>>

\title{Computing MEM seeding probabilities}

@@

The focus of this document is not to solve the true location problem
\textit{per se}, but rather to evaluate whether a solution is correct. At
first sight, whether we can map reads to their true location seems to
depend only on their lengths and on the error rate of the instrument.
However, there is another important factor to consider: most genomes
contain repeated sequences, \textit{i.e.} relatively large subsequences
(from $\sim20$~bp to more than 10~kb) that are present at more than one
location. So we cannot map the fragment with certainty if there exists an
exact copy of the sequence at somewhere else in the genome, because we
cannot know which of the two copies was sequenced.

@@

\subsection{Seeding heuristics}
\label{sec:seedheur}

@@

This definition presents a computational challenge: to know if a given
subsequence is a MEM seed, we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to identify all the MEM seeds from a read, even
for very large genomes\cite{ferragina2000opportunistic,
ferragina2005indexing}. Such algorithms are crucial for the present
discussion, but describing them falls outside the scope of the document.
We refer the interested reader to the relevant
literature\cite{pmid24336412,pmid25399029,pmid23349213,pmid19389736,
li2013aligning}, and here on we will assume that MEM seeds are simply
available at all times. Looking ahead to section~\ref{sec:est}, let
us just say that this can be done with an algorithm called the
\emph{backward search}\cite{ferragina2000opportunistic}.

@@

From now we will make the simplifying assumption that a read with a good
seed is always mapped at the correct location. In other words, we assume
that a good seed is necessary and sufficient for a true positive or
outcome $i.$ In practice this is not always the case, but this assumption
shifts the burden of mapping to the alignment algorithm and allows us
focus on improving the seeding heuristic.

We first explain a detailed strategy to compute the probablity of outcome
$i.$ We then explain how to compute the probability of outcome $iii.$,
\textit{i.e.} that the seeding step returns no hit. Combining these two
quantities allows us to compute the probability of interest, which is that
the read is mapped to an incorrect location or outcome $ii.$ Finally, we
provide a computationally efficient method for estimating the parameters
that are necessary to estimate the probabilities of interest.

@@

\section{Combinatorial construction}

In this section we represent reads as combinatorial ojects. The purpose of
this construction is to give a generative model of reads without good
seeds so that we can later compute their probability of occurrence.

For each potential location including the target, we define the
\emph{match streak} at a given nucleotide as the number of nucleotides
since the last mismatch on the left, or since the beginning of the read.
For instance, in Fig.~\ref{fig:MEM_example}, the match streak of the
target at the ninth nucleotide is $1$ while for the second potential
location, it is $7$.

For instance, the second location in the example above is a hard mask at
the ninth nucleotide. On Fig.~\ref{fig:MEM_example}, one can further see
that at the second nucleotide, all the off-target locations are soft masks
and that at the last nucleotide, the second location is a soft mask.

From the definition, the last nucleotide of every strict on-target MEM
seed is always unmasked. Conversely, an unmasked nucleotide always belongs
to exactly one strict on-target MEM seed. Also, the last nucleotide of
every shared on-target MEM seed is always soft-masked, but a soft-masked
nucleotide does not always belong to a shared on-target MEM seed.

Since hard and soft masks inform us about the positions of on-target
MEM seeds, we construct an alphabet that encodes the masking status of the
nucleotides.

@@

We recode the reads as sequences of letters from the so-called MEM
alphabet $\mathcal{A} = \{\square, |, \uparrow_{/1}, \uparrow_{/2},
\uparrow_{/3} \ldots, \downarrow_{/0}, \downarrow_{/1}, \downarrow_{/2},
\ldots\}$.

The symbols $\downarrow_{/m}$ indicate that the nucleotide is a sequencing
error and $m \geq 0$ is the number of off-target locations that
\emph{match} the nucleotide. Since a sequencing error is always a mismatch
against the target, the symbol $\downarrow_{/0}$ indicates that the
ncleotide is a mismatch against \emph{every} candidate.

The symbols $\uparrow_{/j}$ indicate a change in masking status: the
nucleotide is not masked but its left neighbor is. The index $j \geq 1$ is
the number of nucleotides since the last mismatch or since the beginning
of the read.

All the other nucleotides are represented by the $\square$ symbol.
Finally, the $|$ symbol is appended after the last nucleotide of the read.
It is an obligatory terminator that does not correspond to an actual
nucleotide.

Note that the numbers decorating $\uparrow$ and $\downarrow$ are different
in nature. In the case of the symbol $\downarrow$ it is an information
about the number of sequences matching the read; in the case of the symbol
$\uparrow$ it is the position of the previous sequencing error.
Fig.~\ref{fig:sketch_extended} shows the encoding of the read from
Fig.~\ref{fig:MEM_example} in the MEM alphabet.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_extended.pdf}
\caption{\textbf{The extended MEM encoding}.
The read of Fig.~\ref{fig:MEM_example} is represented in the MEM alphabet.
The arrows departing from those numbers help understand their meaning. The
$\downarrow$ symbol is decorated by the number of locations that match the
nucleotide. The $\uparrow$ symbol is decorated by the number of
nucleotides from the last error or from the beginning of the read. Also
note the presence of the terminator $|$ at the end of the read.}
\label{fig:sketch_extended}
\end{figure}

The MEM alphabet captures the masking status of the nucleotide as follows:
If we denote the number of off-targets as $N$, the symbol
$\downarrow_{/m}$ indicates that the nucleotide has $m$ hard masks and
$N-m$ soft masks. In addition, the symbols $\uparrow_{/j}$ indicate that
the nucleotide is unmasked.

In the MEM alphabet, strict on-target MEM seeds are the longest stretches
of symbols containing $\uparrow$ and not containing $\downarrow$. Indeed,
such a stretch matches the target, it matches the target only because it
contains at least one unmasked nucleotide (marked by $\uparrow$), and it
cannot be extended because it is flanked by sequencing errors or by the
ends of the reads. Note that there is exactly one symbol $\uparrow$ per
strict on-target MEM seed, and therefore that two symbols $\uparrow$ must
be separated by at least one symbol $\downarrow$.

Shared on-target MEM seeds are the longest stretches of $\square$ symbols
flanked by $\downarrow_{/0}$, or by the ends of the read. Indeed, such a
stretch is a MEM seed because it matches the target and it cannot be
extended ($\downarrow_{/0}$ is a mismatch against every sequence). Also,
it cannot be a strict on-target MEM seed because it does not contain the
$\uparrow$ symbol, so it must be shared.

@@

Reads can be viewed as sequences of symbols, or as sequences of
\emph{segments}. For the purpose of constructing reads without good seeds,
the second point of view turns out to be essential.

@@

In line with definition~\ref{def:segment}, the symbols $\uparrow_{/j}$,
$\downarrow_{/m}$ and $|$ are referrred to as \emph{terminators}.
Fig.~\ref{fig:sketch_segment} shows the segment representation of the read
from Fig.~\ref{fig:sketch_extended}.

\begin{figure}[h]
\centering
\includegraphics[scale=.85]{sketch_segments.pdf}
\caption{\textbf{The extended MEM segment encoding}.
The read from Fig.~\ref{fig:sketch_extended} is represented as a sequence
of segments. Each segment is terminated by either $\uparrow_{/j}$, or
$\downarrow_{/m}$ or $|$. The rightmost fragment is the tail. The numbers
below indicate the size of each segment in number of nucleotides.}
\label{fig:sketch_segment}
\end{figure}

We now have a representation of reads as sequences of segments. Only some
of these sequences are possible. For instance, a segment terminated by
$\uparrow$ cannot follow another. The task, however, is not to give a
construction of reads, but a construction of reads without good seeds. For
this, we need to introduce a new set of tools.

\section{The computational strategy}
\label{sec:symbolic}

We turn to the question of translating the extended MEM segment
representation to a construction of reads without good seeds. For this, we
use tools borrowed from analytic combinatorics. We give the minimum amount
of background that is necessary to expose the theory. A more detailed
account of the strategy to compute seeding probabilities with analytic
combinatorics has been published
elsewhere\cite{filion2017analytic,filion2018analytic}. We refer the
readers interested in learning more about analytic combinatorics to the
textbooks\cite{flajolet2009analytic,sedgewick2013introduction}.

@@

The central object of analytic combinatorics is the \emph{generating
function}.

\begin{definition}
\label{def:GF}
Let $\mathcal{A}$ be a set of combinatorial objects characterized by a
size in $\mathbb{N}$ and a weight in $\mathbb{R}^+$. The weighted
generating function of $\mathcal{A}$ is defined as

\begin{equation}
\label{eq:GF1}
A(z) = \sum_{a \in \mathcal{A}} w(a) z^{|a|},
\end{equation}
where $|a|$ and $w(a)$ denote the size and weight of the object $a$,
respectively. Expression (\ref{eq:GF1}) also defines a sequence $(a_k)_{k
\geq 0}$ such that

\begin{equation*}
A(z) = \sum_{k=0}^\infty a_k z^k.
\end{equation*}

By definition $a_k = \sum_{a \in A_k}w(a)$, where $A_k$ is the class of
objects of size $k$ in $\mathcal{A}$. The number $a_k$ is the
total weight of objects of size $k$.
\end{definition}

To give a concrete example, if the probability of the symbol
$\downarrow_{/0}$ is $q$, the weighted generating function of a segment
containing only this symbol is $qz$. In our case, we are interested in the
weighted generating function of reads without good seeds, so that the
coefficient $a_k$ is equal to the probability that a read of size $k$ has
no good seed.

The motivation for introducing weighted generating functions is that
operations on combinatorial objects translate into operations on their
weighted generating functions. If $A(z)$ and $B(z)$ are the weighted
generating functions of two mutually exclusive sets $\mathcal{A}$ and
$\mathcal{B}$, the weighted generating function of $\mathcal{A} \cup
\mathcal{B}$ is $A(z) + B(z)$, as evident from expression (\ref{eq:GF1}).
Size and weight can be defined for pairs of objects in $\mathcal{A} \times
\mathcal{B}$ as $|(a,b)| = |a| + |b|$ and $w(a,b) = w(a)w(b)$. In other
words the sizes are added and the weights are multiplied.  With this
convention, the weighted generating function of the Cartesian product
$\mathcal{A} \times \mathcal{B}$ is $A(z)B(z)$. This simply follows from
expression (\ref{eq:GF1}) and from

\begin{equation*}
A(z)B(z) =
\sum_{a\in \mathcal{A}}w(a)z^{|a|} \sum_{b\in \mathcal{B}}w(b)z^{|b|}
= \sum_{(a,b) \in \mathcal{A} \times \mathcal{B}} w(a)w(b)z^{|a|+|b|}.
\end{equation*}

These two operations are all we need in order to compute the weighted
generating functions of the reads of interest. Addition corresponds to
creating a new family by merging reads from families $\mathcal{A}$ and
$\mathcal{B}$. Multiplication corresponds to creating a new family by
concatenating reads from families $\mathcal{A}$ and $\mathcal{B}$.

\subsection{Analytic representation}

@@

@@

We assume that the target sequence has $N$ duplicates, so that there are
$N$ off-target sequences. We further assume that duplication was
instantaenous and that all $N+1$ sequences diverge independently of each
other at a constant rate. In other words, we ignore the complications due
to the genealogy of the duplication events and we simply assume that at
each position, any given duplicate is identical to the target with
probability $1-\mu$. If it is not, we assume that the duplicate has any of
the remaining three nucleotides with equal probability (\textit{i.e.} each
is found with probability $\mu/3$).

@@

Read errors are always mismatches against the target (because we assume
that the target is the true sequence), and they match each off-target
sequence with probability $\mu/3$. Correct nucleotides are always a match
for the target, and they match each off-target sequence with probability
$1-\mu$.

We are now ready to describe the transfer matrix $M(z)$. We do this entry
by entry, in order of increasing complexity.

@@

A segment terminated by $\uparrow_{/j}$ is the beginning of an on-target
MEM seed of size at least $j$. The seed reaches the next sequencing
error or the end of the read, so the number of $\square$ symbols in the
next segment must be at most $\gamma-j-1$ and it must be terminated by a
$\downarrow$ symbol or by the tail terminator $|$.

On this segment, the matches between the read and the off-target sequences
are irrelevant. The weighted genearting function of each $\square$ symbol
is simply $qz$ ($q$ is the probability that the nucleotide is not an error
and $z$ marks objects of size $1$). The following definition will simplify
the notations.

@@

The weighted generating function of the $\downarrow_{/m}$ terminator is
thus $\omega_m pz$ (the probability of the symbol $\downarrow_{/m}$ is
$\omega_m p$ and $z$ marks objects of size $1$).

@@

The symbol $\downarrow_{/m}$ signifies that the nucleotide has $m$ hard
masks and $N-m$ soft masks. If all the masks disappear before the first
read error, the next terminator will be a $\uparrow$ symbol, otherwise it
will be a $\downarrow$ symbol. We separate the cases based on the
terminator of the segment.

@@

\begin{definition}
The probability that a given off-target sequence contains a mismatch in a
sequence of $i$ error-free nucleotides is
\begin{equation}
\label{eq:xi}
\xi_i = 1-(1-\mu)^i.
\end{equation}
\end{definition}

@@

In the equation above, $q^i$ is the probability that there is no
sequencing error in $i$ nucleotides and $z^i$ marks objects of size $i$.
The fact that the reads have no good seeds imposes $i < \gamma$. Also note
that this expression is the same for all $\downarrow$ symbols; it does not
depend on $m$.

@@

In this case there can be no good seed because the read terminates too
early. We must just enforce the condition that at least one of the $N$
masks survives until the end of the segment in order to exclude the
$\uparrow$ symbol. The weighted generating function is
\begin{equation*}
\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i.
\end{equation*}

@@

In this case, the soft masks do not hide the target: even if one of them
survives until the end of the read, there will be a good seed (shared in
this case). To make sure that the read has no good seed, we must enforce
the condition that at least one hard mask survives until the end of the
segment (which is impossible if $m = 0$). The weighted generating function
is
\begin{equation*}
\sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
\end{equation*}

Finally, the weighted generating function of the tail segments following
segments terminated by $\downarrow_{/m}$ is
\begin{equation}
\label{eq:C}
C_m(z) =
\sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i +
  \sum_{i=\gamma}^\infty \Big(1 - \xi_i^m \Big) (qz)^i.
\end{equation}

@@

In this case, there can be no good seed and we must only exclude the
$\uparrow$ terminator. The weighted generating function is
\begin{equation}
\omega_n pz \sum_{i=0}^{\gamma-1} \Big(1 - \xi_i^N \Big) (qz)^i.
\end{equation}

\subsubsection*{Case 3b: the terminator $\downarrow_{/n}$ comes after the
$\gamma$-th nucleotide}

This case is by far the most convoluted. Since the segment contains at
least $\gamma$ error-free nucleotides, we must ensure that it does not
contain a good seed. This will be the case if any of the two following
conditions is validated: $i)$ at least one hard mask covers all the
error-free nucleotides, or $ii)$ all the hard masks vanish but at least
one soft mask covers the whole segment (including the terminator).

The two conditions are mutually exclusive by construction. They are
graphically represented in the diagram below. The left panel corresponds
to case $i)$ and the panel to case $ii)$. The top row represents the
target, and the bottom rows represent off-target sequences (using the same
symbols as in Fig.~\ref{fig:MEM_example}).
\begin{inset}
\includegraphics{masks.pdf}
\end{inset}

Whenever a hard mask (here the first off-target sequence) covers the
nucleotides as shown in the left panel, there can be no good seed,
irrespective of the value of $\gamma$. The positions marked with a
question mark are irrelevant, they cannot change the fact that there is no
on-target MEM seed. If the hard masks vanish, as in the right panel, then
it all depends on the soft masks. If a soft mask covers the whole segment,
then there can be no good seed, irrespective of the value of $\gamma$.

@@

Condition $ii)$ is more convoluted, so we introduce some further
notations to solve this sub-case.
\begin{definition}
The probability that a given off-target sequence contains a mismatch in a
sequence of $i$ error-free nucleotides followed by an error is
\begin{equation}
\label{eq:eta}
\eta_i = 1-(1-\mu)^i\mu/3.
\end{equation}
\end{definition}

The probability of condition $ii)$ is
\begin{equation*}
\xi_i^m \Big(1 - \eta_i^{N-m} \Big),
\end{equation*}
but we need to break up this term among all the terminators
$\downarrow_{/n}$, ($0 \leq n \leq N$). For this, we decompose the sum on
the number of soft masks that run to the end of the segment, including the
terminator. The probability that there are $r \geq 1$ such soft masks is
\begin{equation*}
{N-m \choose r} (1 - \eta_i)^r \eta_i^{N-m-r}.
\end{equation*}

Each of them matches the nucleotide at the end of the segment, so the
total number of matches is $r$ plus the number of sequences that also
match the last nucleotide, among the remaining $N-m-r$ soft masks and $m$
hard masks.

Let us start with the $N-m-r$ that vanished somewhere in the segment. The
conditional probability that they match the last nucleotide given that
they vanish somewhere in the segment is $\mu/3 \cdot \xi_i / \eta_i$. The
conditional probability that they do not match it is $(1-\mu/3) / \eta_i$.
For the $m$ hard masks, the probability that they match the last
nucleotide given that they vanish somewhere in the segment is simply
$\mu/3$, and the probability that they do not match it is $1-\mu/3$.

The probability that the last nucleotide of the segment matches $n-r$ of
the $N-r$ masks that vanished before the terminator is
\begin{equation*}
\frac{(\mu/3)^{n-r}(1-\mu/3)^{N-n}}{\eta_i^{N-m-r}}
\psi_{i,m,n,r}\text{, where \;}
\psi_{i,m,n,r} = \sum_{q \geq 0}{m \choose q}{N-m-r \choose n-r-q}
\xi_i^{n-r-q}.
\end{equation*}
